---
title: "housing_wage_data"
author: "brian avery"
date: "December 30, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose  

Collate data to make visible the proportion of households in each legislative district that make below what is required to rent a 2 bedroom apartment in their county.  

for [Utahns for Fair Wages](https://www.utahnsforfairwages.org/)  

## Preliminaries

load packages
```{r message=FALSE}
# for dplyr (and maybe ggplot2)
library(tidyverse)

# for kable for better looking html tables at end
library(knitr)
```

## Income data by district  

bring the income data csv files in.

```{r}
house <- read.csv("House_table19c.csv")
senate <- read.csv("Senate_table19c.csv")
```

start cleaning:  
remove `%` signs from percent columns all at once.  
idea from: https://stackoverflow.com/questions/32857873/remove-a-character-from-the-entire-data-frame

```{r}
senate[] <- lapply(senate, gsub, pattern='%', replacement='')
house[] <- lapply(house, gsub, pattern='%', replacement='')
```

convert character values to numbers.  

```{r}
senate[] <- lapply(senate, as.numeric)
house[] <- lapply(house, as.numeric)
```

<b>House data:</b>  
only need first 2 percentage columns, but sorted by rank so districts in different orders.
first, make 2 separate dataframes, 1 for each of the 2 lowest brackets and sort by district.

```{r}
houseL1 <- house[,c("District_L15", "Percent_L15")]
houseL1 <- arrange(houseL1, District_L15)

houseL2 <- house[,c("District_1530", "Percent_1530")]
houseL2 <- arrange(houseL2, District_1530)
```

now combine them together and make a combined column by adding the percentages together.
then only keep the total column and relabel the columns.

```{r}
houseLcomp <- houseL1
houseLcomp$Percent_1530 <- houseL2$Percent_1530
houseLcomp$total_below30 <- houseLcomp$Percent_L15 + houseLcomp$Percent_1530
houseLcomp <- select(houseLcomp, District=District_L15, `%district<30K/yr`=total_below30)
```

<b>Senate data:</b>  
only need first 2 percentage columns, but sorted by rank so districts in different orders.
first, make 2 separate dataframes, 1 for each of the 2 lowest brackets and sort by district.  

```{r}
senateL1 <- senate[,c("District_L15", "Percent_L15")]
senateL1 <- arrange(senateL1, District_L15)

senateL2 <- senate[,c("District_1530", "Percent_1530")]
senateL2 <- arrange(senateL2, District_1530)
```

now combine them together and make a combined column by adding the percentages together.
then only keep the total column and relabel the columns.

```{r}
senateLcomp <- senateL1
senateLcomp$Percent_1530 <- senateL2$Percent_1530
senateLcomp$total_below30 <- senateLcomp$Percent_L15 + senateLcomp$Percent_1530
senateLcomp <- select(senateLcomp, District=District_L15, `%district<30K/yr`=total_below30)
```

## Housing wages and district/county key  

bring in housing wage data and key that links district to county

```{r}
hwage <- read.csv("Housing_Wage_County.csv")
house_counties <- read.csv("house_counties.csv")
senate_counties <- read.csv("senate_counties.csv")
```

join Senate data and make `estimate` column to indicate whether the final percentage is an over or under estimate.

```{r}
senate_j1 <- inner_join(senate_counties, hwage)
senate_j2 <- inner_join(senate_j1, senateLcomp)
senate_j2$estimate <- ifelse(senate_j2$hw_year>30000, "under", "over")
```

join House data and make `estimate` column to indicate whether the final percentage is an over or under estimate.

```{r}
house_j1 <- inner_join(house_counties, hwage)
house_j2 <- inner_join(house_j1, houseLcomp)
house_j2$estimate <- ifelse(house_j2$hw_year>30000, "under", "over")
```

## Final products  

write final dataframes to csv files.

```{r}
write.csv(senate_j2, file="senate_housing_wage.csv", row.names = FALSE)
write.csv(house_j2, file="house_housing_wage.csv", row.names = FALSE)
```

make a prettier final table of House data.  

```{r}
kable(house_j2)
```

make a prettier final table of Senate data.  

```{r}
kable(senate_j2)
```


